---
title: "05_Functions"
author: "Thorsten Brach"
date: "6/2/2017"
output: html_document
---

- helps for the exercises: https://github.com/peterhurford/adv-r-book-solutions

- **The most important thing to understand about R** is that functions are objects in their own right


# Function Components

- The three main components of a function are:
    - body()
    - formals() (list of arguments)
    - environment()
- printing a function shows you these three components, the environment is left out if the function was created in the global environment
- like all R objects, functions in R can have additional attributes (attributes())
- The **srcref** attribute in base R might show you code with comments unlike body() 


```{r}
body(mean)
formals(mean)
environment(mean)
attributes(mean)
```

## Primitive functions

- they are the exception in not having 3 components (formals, body, and environment are NULL), they are implemented in C
- They call C code directly with .Primitive()
- primitive functions are only found in the base package, they can be more efficient (e.g. no copying on modulation), but since they behave differently from other functions the R team avoids them

```{r}
body(sum) #NULL
```

## Exercises

```{r}
# 1.) What functions tell you whether an object is a function or even a primitive function
is.function(mean)
is.primitive(sum)
is.primitive(mean)

# 2.)

### This code makes a list of all functions in the base package.
### 
    objs <- mget(ls("package:base"), inherits = TRUE)
    funs <- Filter(is.function, objs)
### Use it to answer the following questions:
### 
### 1. Which base function has the most arguments?
### 
### 2. How many base functions have no arguments? What's special about those
###    functions.
###    
### 3. How could you adapt the code to find all primitive functions?

# -- 1. Which base function has the most arguments? --
    length(funs) # 1206 functions
# We can use the following to find that the "scan" function has the most arguments:
  which.max(lapply(funs, function(x) length(formals(x))))
# scan
#  908
# The scan function has 22 arguments:
  length(formals(scan))
# [1] 22
# ----

  
# -- 2 How many have no arguments and what is special about them --

funs_0arguments <- funs[sapply(funs, function(x) length(formals(x))==0)]
length(funs_0arguments) #225 have no arguments
# since primitive functions have formals(x) = 0 I assume many of the functions are primitive
sum(sapply(funs_0arguments, is.primitive)) # indeed 183 are primitive
# what with the others
funs_0arguments_notprimitives <- funs_0arguments[!(sapply(funs_0arguments, is.primitive))]
sapply(funs_0arguments_notprimitives, is.function) # of course all true
# so honestly could not figure out what is so special other than that they maybe are in C??

# 3. adapt the code to only find primitives 
  primitive.funs <- Filter(is.primitive, objs)
# There are 181 of them

```

# Lexical scoping

- Scoping are the rules R uses to look up the value of a symbol
- R has two types of scoping:
    - lexical scoping (referring to the computer science term lexing: the process of turning text to pieces that the programming language understands)
    - dynamic scoping (used to save typing during interactive analysis)
- lexical scoping:
    - **looks up symbol values based on how functions were nested when they were created (not when they are called)**. Advantage: the call makes no difference, you can see the look up by looking at the function's definition
    
    
There are four principles behind R's implementation of lexical scoping. 

## 1.) Name Masking

- if a name is not defined within a function, R will look one level up
- not sure why this principle is called name masking but see the last example

```{r}

x <- 2
g <- function() {
y <- 1
c(x, y)
}
g()
rm(x, g)

# example of a closure

j <- function(x) {
        y <- 2
        function() {
                c(x, y)
        }
}
k <- j(23)
k() # the y is saved in the environment of k (environment(k))
y = 4
k() # Maybe this is why it is named Name masking, the k still refers to the y in his environment.
rm(j, k)
```

- btw: closures are functions created by other functions

## 2.) Functions vs. variables

- remember scoping is finding the value of a symbol, the same principles apply regardless of the type of the value to find, i.e. finding functions works just as finding variables
- there is one tweak though, if it is obvious that what you want is a function, R will ignore non-function objects while searching. 
    - see the following example where n is different for the function and the variable (so R scopes for n but knows in the first case it i looking for a function!)
    
```{r}

n <- function(x) x / 2
o <- function() {
n <- 10
n(n)
}
o()

```

## 3.) A fresh start

- every time a function is called, a new environment is created to execute the function. 
   - **Thus a function will not remember values generated in previous runs of the function**

```{r}

j <- function() {
        if (!exists("a")) {
                a <- 1
        } else {
                a <- a + 1
        }
        print(a)
}
j()
j()
```   

## 4.) Dynamic lookup

- lexical scoping determines where to look for a value of a symbol (namely defined by the environments during creation)
- But it is not defined when to look for them, that happens during run of the functions. So the output of a function can depend on objects in environments outside of it (clear). 

- ** For me all these principles are just: scope in the environment of creation, then one up and up and up to the global environment** And you can scope after values that can be functions.
- NB: it is risky when a function depends on values that are in another environment (like the global environment), the function is then not **self-contained**. findGlobals() helps you to find external dependencies of a function.
- but note it is impossible to make a function completely self-contained because R lexical scopes everything, even operators such as +, so you always rely on base R functions

```{r}

f <- function() x + 1
codetools::findGlobals(f)

```   


### Exercises

```{r}
# 1.) What does the following code return? Why? What does each of the three c's mean?

c <- 10
c(c = c) # returns 10 named c. Remember R knows when to look for a function and when for a variable. 

# 2.) For the four principles see headers above
# 3.) What does the following function return, make a prediction first

f <- function(x) {
        f <- function(x) {
                f <- function(x) {
                        x ^ 2
                }
                f(x) + 1
        }
        f(x) * 2
}
f(10)
# I predict 202

```  

# Every operation is a function call

- everything in R is an object, everything that happens is a function call
- this includes **infix** operators such as +
- this includes **control flow** operators such as for
- this includes **subsetting** operators such as {}

```{r}
sapply(1:10, `+`, 3) # uses the value of the object +, which is the function +
sapply(1:10, "+", 3) # match.fun() is used in sapply to find the function named +.

```

# Function arguments

- a function has formal arguments (property of the function) and actual or calling arguments (can vary for each call)
- here you learn how calling arguments are mapped to formal arguments

## Calling functions

- Arguments can be specified by name, partial name, and postion in a function call. 
- you can call a function with a argument list using **do.call**

```{r}
do.call(mean, list(1:10, na.rm = TRUE))
```

## Default and missing arguments

- Because arguments in R are evaluated lazily, you can define default values based on other arguments or even variables in the function body (bad practice)
-the missing function tells you if an argument was supplied (Hadley instead uses default = NULL and then is.null to check if it was provided)
```{r}
g <- function(a = 1, b = a * 2) {
c(a, b)
}
i <- function(a, b) {
c(missing(a), missing(b))
}
i()
```

## Lazy evaluation

- Arguments are only evaluated when they are used
- force() can be used to make arguments be evaluated in any case
- closures are functions created by functions

```{r}
add <- function(x) {
        # he recommends a force(x), or just x here
function(y) x + y
}
adders <- lapply(1:10, add)
adders[[1]](10)
#> [1] 11
adders[[10]](10)
#> [1] 20
```

- remember default values are evaluated inside the function
- an unevaluated argument is called a **promise**

```{r}
f <- function(x = ls()) {
a <- 1
x
}
f()
f(ls()) # global environment
```

- remember if statements are functions, here laziness is handy

```{r}
if (is.null(a)) stop("a is null") # is equal to
!is.null(a) || stop("a is null") # thanks to lazy evaluation

```

## The ... argument

- matches all arguments not otherwise matched
- useful for example to collect arguments to call another function
- ... is often used with S3 generic functions to allow individual methods to be more flexible
- advantage: makes function flexible (like plot) but more complicated to understand.  
- use list to capture ... in an easier to work with form

```{r}
f <- function(...) {
list(...)
# or names(list(...))
}
f(a = 1, b = 2)
```
- ... invites for unnoticed typos

### Exercises

```{r}
# 1. Clarify the following list of odd function calls (not done)

# 2. What does this function return? Why? Whcih principle does it illustrate

f1 <- function(x = {y <- 1; 2}, y = 0) {
x + y
}
f1()
# it returns 3. y is default to 0, but then x is lazily evaluated within the function which assigns a new value to y. 
# so I would say it illustrates lazy evaluation

# 3. What does this function return? Why? Whcih principle does it illustrate
f2 <- function(x = z) {
z <- 100
x
}
f2()
# returns 100, again illustrating lazy evaluation
```
