---
title: "09_Functional programming"
author: "Thorsten Brach"
date: "2/3/2017"
output: html_document
---

- for exercises check: <https://github.com/peterhurford/adv-r-book-solutions>

# Functional programming (FP) basics

- FP means you create and manipulate functions. 
- the three bilding blocks of FP are
    - anonymous functions
    - closures (functions written by functions)
    - lists of functions
- in FP:
    - build easy to understand building blocks
    - combine them into complex structures applying each block with confidence
- remember the Do not repeat yourself dogma, so always think about removing repetition
    
```{r}
library(pryr)
```

- lapply is a functional, because it uses a function as input and puts out a vector.
- remember to use [] when applying lapply to a data frame
- this is the idea of function composition (building blocks)

```{r}
# e.g. df[] <- lapply(df, fix_missing)
```


# Anonymous functions

- keep functions anonymous only if it fits one line and does not use {}


# Closures

- Closures are functions written by functions.
- The enclosing environmnet of a closure (I.e. environment(closure)) is the execution environment of the function that created it. 


```{r}
power <- function(exponent) {
print(environment())
function(x) x ^ exponent
}

zero <- power(0)
environment(zero)
```


# Mutable state

- Remember that <<- is a way to assign values to parent (path up) environments. When using in a closure it allows to make changes in the enclosing environment of the closure, which is the parent environment of the execution environment. 


```{r}
new_counter <- function() {
i <- 0
function() {
i <<- i + 1
i
}
}
```


# Moving lists of functions to the global environment


```{r}
simple_tag <- function(tag) {
force(tag)
function(...) {
paste0("<", tag, ">", paste0(...), "</", tag, ">")
}
}
tags <- c("p", "b", "i")
html <- lapply(setNames(tags, tags), simple_tag)
# Why do you have to setNames first?
html2 <- lapply(tags, simple_tag)
# because names(html2) is NULL
names(html2) <- tags # then it would be the same
```

- so the deal of setNames was that lapply keeps the names of the list
- to get the list to the global environment you have three options
    - you do not want to have the functions always in the global environment because they have dangerous names that could clash
    
    
```{r}
# option 1 with
with(html, p("HOrst"))

# option 2 attach and detach
attach(html)
p("Horst")
detach(html)

# option 3: list2env
list2env(html, environment())
```

- keep in mind the list2env function, or all three options in how you quickly get list variables to the global environment!



# Case study: numerical integration


- integration by midpoint means with rectangles
- we want to integrate sin(x) from 0 to pi, the result is 2

```{r}
# - see what we are integrating on -
x = seq(0, pi, by = 0.01)
plot(x, sin(x))
integrate(sin, 0, pi)
# --

midpoint <- function(f, a, b) {
(b - a) * f((a + b) / 2)
}
# you want to integrate from a to b, so here just the x stretch times the value at the middle of the curve

trapezoid <- function(f, a, b) {
(b - a) / 2 * (f(a) + f(b))
}
midpoint(sin, 0, pi)

```

- you clearly need to break up the range into smaller pieces


```{r}
midpoint_composite <- function(f, a, b, n = 10) {
points <- seq(a, b, length = n + 1)
h <- (b - a) / n
area <- 0
for (i in seq_len(n)) {
area <- area + h * f((points[i] + points[i + 1]) / 2)
}
area
}

trapezoid_composite <- function(f, a, b, n = 10) {
points <- seq(a, b, length = n + 1)
h <- (b - a) / n
area <- 0
for (i in seq_len(n)) {
area <- area + h / 2 * (f(points[i]) + f(points[i + 1]))
}
area
}

midpoint_composite(sin, 0, pi, n = 10)
```

- already very good!
- notice that there is a lot of repetition between trapezoid and midpoint_composite. You could do it more general


```{r}
composite <- function(f, a, b, n = 10, rule) {
points <- seq(a, b, length = n + 1)
area <- 0
for (i in seq_len(n)) {
area <- area + rule(f, points[i], points[i + 1])
}
area
}
composite(sin, 0, pi, n = 10, rule = midpoint)
```

- I thought the part on more complicated rules is not that exciting.